---
title: "Tea Time Theme Time Term Matching"
author: "Michael Pawlus"
layout: post
published: no
status: process
draft: no
---

### A quick code snippet to find matching terms between song lyrics

I love [BBC 6 Music](https://www.bbc.co.uk/6music). On my morning train rides into LA I like to listen to the [Radcliffe and Maconie Show](http://www.bbc.co.uk/programmes/b0100rp6). The hosts play a game at the end of their show where they play three tracks and ask listeners to guess what connects all three tracks.

I have yet to win this game which has prompted me to try to see if I can gain an advantage by analyzing some data about the the themes submitted. I have noticed that since the start of 2018, the theme relies on an exact term or phrase match among the lyrics 17% of the time.

```{r}
library(tidyverse)

# read in the data
t <- read_csv("https://raw.githubusercontent.com/michaelpawlus/michaelpawlus.github.io/master/images/tttt1.csv")

# filter to one link per row, removing song 2 and 3
tr <- t %>% filter(col == "sa1")

# filter to just the rows were links are based on a term or phrase in the lyrics
tl <- t %>% filter(link_type == "Term", link_location == "Lyrics", col == "sa1")

# plot the proportion of cases were the link is a term or phrase in the lyrics
t %>%
select(index, link_location, link_type, col) %>%
filter(col == "sa1") %>%
  select(-col) %>%
  ggplot(aes(link_type)) + 
  geom_bar(aes(fill = link_location)) + 
  annotate("text", x = 'Term', y = 5, label = paste0(round(nrow(tl)/nrow(tr),2)*100,"%"), fontface = 2, size = 6) +
  labs(y = NULL, x = "Link Type") + 
  guides(fill=guide_legend(title="Link Location"))
```

In these cases, these were the links.

```{r}
tl %>% select(Link)
```

Using what I have recently learned from the excellent [Tidy Text Mining](https://www.tidytextmining.com/)  and really helpful [package](https://github.com/JosiahParry/geniusR) that pulls in lyrics from the site [Genius](https://genius.com/), I was able to write a really quick script to get the lyrics for two songs and look for common terms between them. Here is an example to find one of the links above:

```{r}
library(tidytext)
library(geniusR)

song1 <- genius_lyrics(artist = "Bill Withers", song = "Ain't No Sunshine") %>%
  unnest_tokens(trigram, text, token = "ngrams", n = 3) %>%
  separate(trigram, c("word1", "word2", "word3"), sep = " ") %>%
  count(word1, word2, word3, sort = TRUE)

song2 <- genius_lyrics(artist = "Prince", song = "I Could Never Take The Place of Your Man") %>%
  unnest_tokens(trigram, text, token = "ngrams", n = 3) %>%
  separate(trigram, c("word1", "word2", "word3"), sep = " ") %>%
  count(word1, word2, word3, sort = TRUE)

song1 %>%
  inner_join(song2, by = c("word1","word2","word3")) %>%
  count(word1, word2, word3, sort = TRUE)
```

My next steps is to wrap this all in a function and add a check that first looks for matching trigrams and then finding none looks for bigrams and failing that lists all matching terms ommitting stop words. I will post that update soon.

As of this post, I still have yet to win Tea Time Theme Time (#TTTT)